% Cover letter for submission to Springer Nature / Foundations of Physics
% Compile with: pdflatex cover_letter.tex
\documentclass[11pt]{letter}

\usepackage[margin=1in]{geometry}
\usepackage[hidelinks]{hyperref}

% --- Editable fields -------------------------------------------------
\newcommand{\journal}{Foundations of Physics}
\newcommand{\manuscripttitle}{Seam Geometry of Bohmian Mechanics: Geodesics, Foliations, and Nodal Singularities}

\signature{Lars R\"onnb\"ack}
\address{%
Department of Computer and Systems Sciences\\
Stockholm University\\
Stockholm, Sweden\\
\href{https://orcid.org/0000-0002-8810-7587}{ORCID: 0000-0002-8810-7587}\\
\href{mailto:lars@uptochange.com}{lars@uptochange.com}%
}

% Set the recipient address as appropriate for the journal/submission portal.
\begin{document}
\begin{letter}{%
Editorial Office\\
\journal\\
Springer Nature
}

\date{\today}

\opening{Dear Editors of \journal,}

Please consider the enclosed manuscript entitled ``\manuscripttitle'' for publication in \journal.

The paper develops a ``seam'' (scalar-first) geometric reframing of the Bohm--Madelung formulation of quantum mechanics, with the amplitude seam $s=\log R$ encoding the quantum potential via $Q=-(s''+(s')^2)$ in the stationary 1D convention used. It also introduces a quadratic differential $w=(\partial_z^2\log\psi)\,dz^2$ (in a Wirtinger-derivative sense) and compares its induced horizontal direction field to probability-current streamlines. For the Gaussian vortex state $\psi=ze^{-|z|^2/2}$ the paper proves exact direction-field alignment away from the node, and it provides an asymptotic ``vortex dominance'' result near isolated simple zeros under mild local factorisation hypotheses, together with explicit limitations via a two-node non-example and reproducible numerical tests.

This manuscript is original, has not been published previously, and is not under consideration for publication elsewhere.

\textbf{Funding.} The author did not receive support from any organization for the submitted work.

\textbf{Competing interests.} The author has no relevant financial or non-financial interests to disclose.

The manuscript includes a Code Availability statement with a public repository link for figure-generation scripts and numerical checks.

Thank you for your consideration.

\closing{Sincerely,}

\end{letter}
\end{document}
